Likelihoods for a general class of ARGs under the SMC

### 2025-04-23 review-responses.tex

\begin{point}{\revref} % 24:
 evol biol $\to$ genetics?
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 27:
 ancestral processes $\to$ genetic processes?
\end{point}

\reply{
We would prefer to keep ancestral processes here as ARGs contain
information about, for example, population structure which is not
a genetic process, but something that we gains insights into
through genetics.
}

@@

\begin{point}{\revref} % 82:
 heuristic methods
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 84:
 likelihood of a dataset? under a model (to make it more general)
\end{point}

\reply{
We would prefer to keep this as it is to make it clear that it
is the ARGs themselves that we want to compute the likelihood of.
}

@@

\begin{point}{\revref}
I would appreciate if you here make it clear which information is missing in these different tools. Sorry for nitpicking.
\end{point}

\reply{
This is quite a subtle point which we feel would distract from the main
message here by expanding. We have attempted to clarify by stating
that a sequence of local trees does not have enough information to
reconstruct a full ARG (required for likelihood calculation under KYF)
and referred to Wong et al for the details, where there is a full
section dedicated to the correspondence between an  ARG and the
sequence of local trees.
}

### 2025-04-23 paper.tex
Address some more points

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the
details of recombination events, and heuristic methods (in different ways)
omit these details. \revpoint{2}{8}
In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events.

There are substantial advantages to the statistically rigorous sampling
approaches of ARGweaver and SINGER.
In particular, they tend to be more accurate (at the
scale at which they can be compared),
and sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty
around estimates. \revpoint{2}{9}
On the other hand, heuristic methods have the potential
to use the greater amount of information in large datasets,
capturing fine-scale details about the recent past.
% thus possibly producing even more accurate estimates
% (although this has not, to our knowledge, been tested).
Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still
only feasible for hundreds of samples.  \revpoint{2}{1}
Meanwhile, tsinfer, ARG-Needle and Threads have
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigour in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g., MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the
% goodness of fit for model estimates.

A key problem facing the field is that the ARGs estimated by heuristic large-scale \revpoint{2}{10}
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. \revpoint{2}{11}
Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probability to each recombination and common
ancestor event, and to the inter-event  waiting times.
Although the KYF formulation is a natural and elegant way to describe
the likelihood of an ARG, the high level of detail about the underlying
events that is required is not present
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation. \revpoint{2}{12}
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021}
by default outputs a more complete and efficient representation of
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

>>>>

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the
details of recombination events, and heuristic methods
omit these details in different ways. \revpoint{2}{8}
The significant progress made in
scalability by these heuristic methods is in part \emph{because} they don't
attempt to precisely reconstruct recombination events.

There are substantial advantages to the statistically rigorous sampling
approaches of ARGweaver and SINGER.
In particular, they tend to be more accurate (at the
scale at which they can be compared),
and sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty
around point estimates. \revpoint{2}{9}
On the other hand, heuristic methods have the potential
to use the greater amount of information in large datasets,
capturing fine-scale details about the recent past.
% thus possibly producing even more accurate estimates
% (although this has not, to our knowledge, been tested).
Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still
only feasible for hundreds of samples.  \revpoint{2}{1}
Meanwhile, tsinfer, ARG-Needle and Threads have
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigour in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g., MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the
% goodness of fit for model estimates.

A key problem facing the field is that the ARGs estimated by heuristic \revpoint{2}{10}
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. \revpoint{2}{11}
Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probability to each recombination and common
ancestor event, and to the inter-event  waiting times.
Although the KYF formulation is a natural and elegant way to describe
the likelihood of an ARG, the high level of detail about the underlying
events that is required is not present
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail to reconstruct
the ``full ARG'' required for likelihood
computation~\citep{wong_general_2023}. \revpoint{2}{12}
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021}
by default outputs a more complete and efficient representation of
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.
